---
title: 'Homework II: SVD analysis & Life Tables'
author: "Inès Dardouri, Mohamed Boudokhane and Lycia Fezzoua"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_notebook:
    code_folding: none
    number_sections: yes
    toc: yes
  html_document:
    fig_caption: yes
    keep_md: yes
    number_sections: yes
    toc: yes
subtitle: 'Deadline: 2020-05-20'
params:
  country: France
  country_code: fr_t
  dirpath: ../LIFE_TABLES/lt_
  timecourse: 1945:2015
---


---

# Objectives

This notebook aims at

  - working with **tables** (`data.frames`, `tibbles`, `data.tables`, ...) using `dplyr` or any other query language (as provided for example by `data.table`)
  - visualizing demographic data as provided by [Human Mortality Database organization](https://www.mortality.org).
  - using PCA and other matrix oriented methods (CCA) to explore a multivariate datasets (lifetables may be considered as multivariate datasets)

# Life tables data (ETL)

We investigate life tables describing countries from Western Europe (France, Great Britain --actually England and Wales--, Italy, the Netherlands, Spain, and Sweden) and the United States.

We load the one-year lifetables for female, male and whole population for the different countries.


```{r, echo=FALSE, eval=FALSE}
# for debugging
# params should be initialized from YAML header
params<- list(
    timecourse= 1945:2015,
    dirpath= 'LIFE_TABLES/lt_',
    country_code= 'fr_c',
    country= 'France')
```

```{r, echo=FALSE}
timecourse <- eval(rlang::parse_expr(params$timecourse))
```

```{r tidyverse, echo=FALSE, message=FALSE, warning=FALSE}
pacman::p_load(tidyverse)
pacman::p_load(plotly)
pacman::p_load(foreach)
pacman::p_load(iterators)
pacman::p_load(DT)
pacman::p_load(ade4)
pacman::p_load(magrittr)
pacman::p_load(FactoMineR)
pacman::p_load(factoextra)
pacman::p_load(FactoInvestigate)
pacman::p_load(ggfortify)
pacman::p_load(plyr)
pacman::p_load(viridis)
pacman::p_load(CCA)
pacman::p_load(vegan)
pacman::p_load(corrplot)
pacman::p_load(demography)
old_theme <-theme_set(theme_dark(base_size=9,
                                 base_family = "Helvetica"))
knitr::opts_chunk$set(eval=TRUE,
  echo=FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  autodep = TRUE,
  tidy = FALSE)
```


```{r, echo=FALSE}
country_code <- list(fr_t='FRATNP',
                     fr_c='FRACNP',
                     be='BEL',
                     gb_t='GBRTENW',
                     gb_c='GBRCENW',
                     nl='NLD',
                     it='ITA',
                     swe='SWE',
                     sp='ESP',
                     us='USA',
                     ch='CHE',
                     jp='JPN')
countries <- c( 'fr_t', 'gb_t',  'nl', 'it', 'sp', 'swe', 'us', 'ch', 'jp')  
country_names <- list(fr_t='France',     # total population
                     fr_c='France',      # civilian population
                     be='Belgium',
                     gb_t='England & Wales',    # total population
                     gb_c='England & Wales',    # civilian population
                     nl='Netherlands',
                     it='Italy',
                     swe='Sweden',
                     sp='Spain',
                     us='USA',
                     ch='Switzerland',
                     jp='Japan')
gender_names <- list('b'='Both',
                     'f'='Female',
                     'm'='Male')
 
```


```{r load_tabble, warning=FALSE, message=FALSE, results="hide"}
dirpath <- 'LIFE_TABLES/lt_'    
suffix <- 'ltper_1x1.txt'
mid <- 'ltper_1x1/'
foreach(country=iter(countries), .combine = c) %do% {
  purrr::map2_chr(.x = c("both/b", "female/f", "male/m"),
                  .y = c("b", "f", "m"),
                  .f = ~ stringr::str_c(dirpath,
                                  .x ,
                                  mid,
                                  country_code[[country]],
                                  ".",
                                  .y,
                                  suffix))
  }  -> file_paths
df_names <- as.character(outer(c("b", "f", "m"),
                  countries,
                  FUN = function(s, c) stringr::str_c("data", s, c, sep="_")))
out <- purrr::map(file_paths,
                  ~ readr::read_delim(., delim = " ",
                                      skip = 2,
                                      col_names = TRUE))
names(out) <-  df_names
purrr::map2(.x = out,
            .y=stringr::str_split(df_names, pattern = 'data_[m,f,b]_', simplify = TRUE)[,2],
            ~ mutate(.x, Country=country_names[[.y]])) -> out
purrr::map2(.x = out,
            .y =stringr::str_split(df_names, pattern = '_', simplify = TRUE)[,2],
            ~ mutate(.x, Gender=gender_names[[.y]])) -> out
list2env(x = out, envir = environment())
```
```{r inspect_tables,results="hide"}
out['data_b_nl'] %>%
  glimpse()
```


The meaning of the different columns:

`mx`: Central death rate between ages x and x+n where n=1, 4, 5, or ∞ (open age interval)

`qx`: Probability of death between ages x and x+n

`ax`: Average length of survival between ages x and x+n for persons dying in the interval

`lx`: Number of survivors at exact age x, assuming l(0) = 100,000

`dx`: Number of deaths between ages x and x+n

`ex`:: Life expectancy at exact age x (in years)


But some of the columns need retyping:

- `Year`: should be integer
- `Age`:  needs some cleaning, after cleaning it should be typed as integer
- `Lx`: should be integer
- `Tx`: should be integer
- Other columns should be considered as floating point numbers (`numeric`)

```{r results="hide"}
retype <- function(data){
  data %>%
    dplyr::rename_all(stringr::str_trim) ->
    data
  double_cols <-  c("mx", "qx", "ax", "ex")
  integer_cols <- setdiff(names(data), c(double_cols, "Country", "Gender"))
  data %>%
    dplyr::mutate_at(double_cols, as.numeric) %>%
    dplyr::mutate_at(integer_cols, as.integer) %>%
    dplyr::mutate_at(vars(Country,Gender), forcats::as_factor) %>%
    tidyr::drop_na(Age) -> data
  data
}
out %>%
  purrr::map(retype) -> out
list2env(x = out, envir = environment())
```

```{r results="hide"}
count_na <- function(data){
  data %>%
    dplyr::summarise_all(.funs = ~sum(is.na(.)))-> res
  res
}
life_table <- ldply(out, data.frame) 
life_table
```


---

- After ETL processing, we obtain a universal table encompassing
all data available  in the files located in `LIFE_TABLES` directory.
Henceforth, the universal table is named `life_table`, its schema is
the following.

| Column Name | Column Type |
|:------------|:------------|
|  Year       | integer     |
|  Age        | integer     |
|  mx         | double      |
|  qx         | double      |
|  ax         | double      |
|  lx         | integer     |
|  dx         | integer     |
|  Lx         | integer     |
|  Tx         | integer     |
|  ex         | double      |
|  Country    | factor      |
|  Gender     | factor      |

Coercion introduces a subtantial number of NA  warnings.
Preliminary inspection of the data suggests that coercion problems
orginate from column `Age`: `110+` cannot be coerced to an integer value.
We discard corresponding rows using `tidyr::drop_na(Age)`.

```{r result="hide"}
out %>%
  purrr::map(~ sum(is.na(.))) %>%
  as.data.frame() %>%
  pivot_longer(cols= names(.), names_to="Data frame", values_to="# NAs")
data_b_nl  %>%
  summarise_all(~ sum(is.na(.)))
```

# Western countries in 1948

* **Visualization of the central death rates of all Countries at all ages for year 1948 :**

```{r, results="markup", fig.width=8, fig.height=5}
p <- life_table %>%
  filter(Year == '1948', Gender == 'Both') %>%
  ggplot(mapping=aes(x=Age, y=mx)) +
  geom_line(span=.5,aes(colour = forcats::as_factor(Country)), size = .2) +
  scale_y_log10() +
  xlab("Age")+
  ylab("Death rates") +
  labs(colour = "Countries") +
  ggtitle("Central death rates at all ages for year 1948") +
  theme_bw()
print(p)
```

We notice that the death rates for new borns are much higher for Italy and Spain than for the rest of the european countries and for the USA. This difference is still noticable for infant mortality. But for the adults, the death rates are pretty much the same for all countries. The difference for young people's mortality could be explained by the different economic and health conditions at that time between the different countries. 
  
* **Visualization of ratios between central death rates in European countries and central death rates in the USA in 1948 :** 

```{r fig.width=10, fig.height=3}
ratio_us_year <- function(tab,year){
  us_year <- data_b_us %>%
    filter(Gender == 'Both' , Year == year) 
  
  tab %>%
    filter(Gender == 'Both' ,Year == year) %>%
    mutate(ratio_us = mx/us_year$mx) -> tab
  return(tab)
}
out_us_48 <- purrr::map(out, ratio_us_year, 1948)
p <- out_us_48 %>%
  ldply( data.frame) %>%
  filter(Gender =='Both') %>%
  ggplot(mapping=aes(x=Age, y=ratio_us)) +
  geom_smooth( aes(colour = forcats::as_factor(Country)), method="loess", se=FALSE) +
  scale_y_log10() +
  xlab("Age") +
  ylab("Death rates") +
  labs(colour = "Countries") +
  ggtitle("Ratios between central death rates in European countries and central death rates in the USA in 1948") +
  theme_bw()
print(p)
```


We can see that the ratio between central death rate in Netherlands and central death rate in the USA is less than 1 for almost all the ages except the oldest ones, which means that the central death rate in Netherlands is lower than central death rate in the USA in 1948. But we can also see that this ratio is greater than 1 for almost all the other European countries (except Sweden), which means that the central death rate in the majority of the European countries is higher than the central death rate in the USA in 1948, especially for France and Spain. This difference could be explained by the health conditions and the financial situation of the two continents at that time.
  
* **Visualization of ratios between central death rates in European countries and central death rates in the USA in 1965 :** 
```{r fig.width=10, fig.height=3}
out_us_65 <- purrr::map(out, ratio_us_year,1965)
p <- out_us_65 %>%
  ldply( data.frame) %>%
  filter(Gender =='Both') %>%
  ggplot(mapping=aes(x=Age, y=ratio_us)) +
  geom_smooth(aes(colour = forcats::as_factor(Country)), method="loess", se=FALSE) +
  scale_y_log10() +
  xlab("Age") +
  ylab("Death rates") +
  labs(colour = "Countries") +
  ggtitle("Ratios between central death rates in European countries and central death rates in the USA in 1965") +
  theme_bw()
print(p)
```
When we plot the same ratio but almost 20 years later, we can see the evolution of European countries' central death rates.
Indeed, the ratio becomes lower than 1 for the middle ages, which means central death rates of European countries are approching USA's central death rates, even if it's still not the case at the extremities' ages. So, in 1965 Europe started regaining the health gap between her and the USA, comparing to what it was 20 years earlier
  
* **Visualization of ratios between central death rates in European countries and central death rates in the USA in 2015 :** 
```{r fig.width=10, fig.height=3}
out_us_15 <- purrr::map(out, ratio_us_year,2015)
p <- out_us_15 %>%
  ldply( data.frame) %>%
  filter(Gender =='Both') %>%
  ggplot(mapping=aes(x=Age, y=ratio_us)) +
  geom_smooth(aes(colour = forcats::as_factor(Country)) ,method="loess", se=FALSE ) +
  scale_y_log10() +
  xlab("Age") +
  ylab("Death rates") +
  labs(colour = "Countries") +
  ggtitle("Ratios between central death rates in European countries and central death rates in the USA in 2015") +
  theme_bw()
print(p)
```

And if we study again the same ratio but for the year 2015, we clearly see an inversion of the 1948's curve for most of the European countries. Europe's death rates are now much lower than USA death rates for ages from 0 to almost 75. In fact, the economical situation of the Europe was remarkably increasing since the end of the WWII and therefore its health situation. We also have to underline the fact that each country has its own health politics which can be more or less in favor of the population. But we also notice that the USA have always had lower death rates for the oldest ages than European countries, since the end of WWII. This may be explained by the fact that the USA care more about their elderly than about the rest of the population, or simply because more Europeans reach old ages then die quickly.  


# Death rates evolution since WW II

* **Visualization of mortality quotients (column `mx`) for both genders as a function of `Age` for  years `1946, 1956, ...` up to `2016` .**


```{r fig.width=8, fig.height=5}
after_ww_II <- seq(1946,2016, by = 10)
p <- life_table %>%
  filter(Year %in% after_ww_II, Gender != 'Both', Country %in% c('USA', 'Netherlands')) %>%
  ggplot(mapping=aes(x=Age, y=mx)) +
  geom_line(method="loess", se=FALSE, span=.5,aes(colour = forcats::as_factor(Year)), size = .2) +
  scale_y_log10() +
  xlab("Age")+
  ylab("Mortality quotients") +
  labs(colour = "Years") +
  facet_grid(vars(Gender),vars(Country)) +
  ggtitle("Mortality quotients for both genders as a function of Age since after WWII") +
  theme_bw()
print(p)
```



+ Concerning the plot below:

We notice that the mortality quotients of young people in 1946 is smaller in the USA than in all the European countries.
This is certainly due to the fact that the USA didn't suffer a lot from human loss during the WWII, unlike the European countries.


We modify our dataframe so it has the following schema:


| Column Name | Column Type |
|:------------|:------------|
|  Year       | integer     |
|  Age        | integer     |
|  mx         | double      |
|  mx.ref_year| double      |
|  Country    | factor      |
|  Gender     | factor      |

where `(Country, Year, Age, Gender)` serves as a _primary key_,
`mx` denotes the central death rate at `Age` for `Year` and `Gender` in `Country`
whereas `mx_ref_year` denotes central death rate at `Age` for argument `reference_year`
in `Country` for `Gender`.


```{r, results="hide"}
ratio_mortality_rates <- function(df,
                                  reference_year=1946,
                                  target_years=seq(1946, 2016, 10)){
  
  dplyr::filter(df, Year %in% target_years) %>%
  dplyr::inner_join(y=df[df$Year==1946, c("Age","Country", "mx")], by=c("Age","Country")) %>%
  dplyr::filter(Age <90) -> df
  df <- select (df,-c(qx,ax,lx,dx,Lx,Tx,ex ))
  df
}
ratio_mortality_rates(life_table)
```


* **Visualization of the ratio  $m_{x,t}/m_{x, 1946}$ for ages $x \in 1, \ldots, 90$ and year $t$ for $t \in 1946, \ldots, 2016$ where $m_{x,t}$ is the central death rate at age $x$ during year  $t$ :**

```{r}
q <- dplyr::filter(ratio_mortality_rates(data_b_nl), FALSE) %>%
  ggplot(aes(x=Age, y=mx.x/mx.y)) +
  geom_smooth(method="loess",
              se= FALSE,
              size =.5,
              span= .2) +
  facet_wrap(vars(Year)) +
  ylab("Ratio of mortality rates, reference Year 1946") +
  labs(colour="Year", linetype="Year") +
  theme_bw()
```

```{r}
(q  %+%
  ratio_mortality_rates(data_m_nl) +
  ggtitle("Variation of mortality quotient with respect to Y=1946, Males in Netherlands"))
```



+ We handle both genders and countries `Spain`, `Italy`, `France`, `England & Wales`, `USA`, `Sweden`, `Netherlands`.
  
```{r}
(q  %+% 
  ratio_mortality_rates(filter(life_table, Country %in% c("Netherlands","France", "USA"))) +
   facet_grid(vars(Country),vars(Year)) +
  ggtitle("Variation of mortality quotient with respect to Y=1946, both genders"))
```
  
But as we did since the beginning, we concentrate on the comparison between te USA and the Netherlands.

```{r, fig.height=8, fig.width=12}
half_full_data <- life_table %>%
  dplyr::filter(Country %in% c('Netherlands','USA'),Gender == 'Both')
k <- dplyr::filter(ratio_mortality_rates(half_full_data), FALSE)%>%
  ggplot(aes(x=Age, y=mx.x/mx.y)) +
  geom_smooth(se= FALSE,
              size =.5,
              span= .2) +
  facet_wrap(vars(Year)) +
  ylab("Ratio of mortality rates, reference Year 1946") +
  labs(colour="Year", linetype="Year") +
  scale_colour_brewer()
(k  %+% 
  ratio_mortality_rates(half_full_data) +
   facet_grid(vars(Country),vars(Year)) +
  ggtitle("Variation of mortality quotient with respect to Y=1946, both genders")) +
  theme_bw()
```

In the USA, the ratio of mortality rates between all the years after 1946 and the year 1946 has always been under 1 for all ages, which meens that since 1946 people die less in the USA than in 1946. Whereas in the Netherlands, this ratio has been higher than in the USA for all years and especially for the older ages.
We also notice a difference for the new borns between the two countries : the ratio is twice higher for the USA in 1956 than in the Netherlands, which means that the mortality rate in 1946 was much higher in the Netherlands compared to the other years, whereas the difference is smaller for the USA between 1946 and the other years.
The ratio becomes higher in the USA than in the Netherlands for the age 25 since 2006.

# Trends

* **Visualization of mortality quotients at ages $0, 1, 5$  as a function of time, facetted by Gender and Country :**

```{r fig.width=10, fig.height=10}
ages <- c(0, 1, 5)
mx_by_age <- life_table %>%
  filter(Age %in% ages, Gender != 'Both', Country %in% c('France','Italy', 'Spain', 'Netherlands','USA') ) %>%
  ggplot(mapping=aes(x=Year, y=mx)) +
  geom_line(size = 0.4, aes(colour =  forcats::as_factor(Age))) +
  facet_grid(vars(Country), vars(Gender)) +
  scale_y_log10() +
  ylab("mortality quotients")+
  labs(color = 'Age') +
  theme_bw()
mx_by_age
```

New borns die more than children of ages 1 and 5 for both genders and for all countries, and children of age 5 die the less.
We don't notice any difference between the mortality quotients of the two genders, for all the countries.
We can see some noticable peaks for the Netherlands corresponding to the two world wars, that don't appear on the US plot, for the 3 different ages.
Also, for both the USA and our European country, the mortality quotients were obviously higher 100 years ago than what they are nowadays. 


* **Visualization of mortality quotients at ages $15, 20, 40, 60$  as a function of time, facetted by `Gender` and `Country` :**

```{r, fig.width=8, fig.height=16}
ages_2 <- c(15,20,40,60)
mx_by_age_2 <- life_table %>%
  filter(Age %in% ages_2, Gender != 'Both', Country %in% c('France','Italy', 'Spain', 'Netherlands','USA')) %>%
  ggplot(mapping=aes(x=Year, y=mx)) +
  geom_line(size = 0.4, aes(colour =  forcats::as_factor(Age))) +
  facet_grid(vars(Country), vars(Gender)) +
  scale_y_log10() +
  ylab("Mortality quotients")+
  labs(color = 'Age') +
  ggtitle("Variation of mortality quotient at ages 15, 20, 40 and 60 as a function of time") +
  theme_bw()
mx_by_age_2
```



We observe that the mortality quotients for male are higher than for women for ages 15 to 60 for both countries.
We can also see that the Netherlands peaks are still observable and they are even sharper, for the same years (WWI and WWII).
Also, for both the USA and our European country, the mortality quotients were obviously higher 100 years ago than what they are nowadays. 


# Rearrangement

- From our dataframe `life_table`, we then compute another dataframe  called `life_table_pivot`
with primary key `Country`, `Gender` and `Year`, with a column for each `Age` from `0` up to `110`.
For each age column, the entry should be the central death rate
at the age defined by column, for `Country`, `Gender`  and `Year` identifying the row.

The resulting schema looks like:

| Column Name | Type    |
|:------------|:--------|
| Country     | factor  |
| Gender      | factor  |
| Year        | integer |
| `0`         | double  |
| `1`         | double  |
| `2`         | double  |
| `3`         | double  |
| $\vdots$    | $\vdots$|


- Using this new dataframe, we compute life expectancy at birth for each Country, Gender and Year :

```{r results="hide"}
life_table_pivot <- life_table %>%
  select(Country, Gender, Year, Age, mx) %>%
  tidyr::pivot_wider(names_from = "Age", values_from = "mx")
life_table_pivot_log <- life_table %>%
  select(Country, Gender, Year, Age, mx) %>%
  dplyr::mutate(mx=log(mx)) %>%
  tidyr::pivot_wider(names_from = "Age", values_from = "mx")
ex <- function(mx){
  sum(cumprod(1 - mx))
}
life_table_pivot %>%
    select(-c("Country","Gender","Year")) %>%
    apply(1, ex) -> life_table_pivot$life_ex 
life_table_pivot  
```


# Life expectancy

- We then calculate the residual life expectancy corresponding to the vector of mortality quotients and a given age.

\[
ex = \sum_{} \prod_{} 1-mx
\]


```{r}
ex <- function(mx){
  sum(cumprod(1 - mx))
}
```

```{r}
conv_to_A <- function(data){
  data %>% 
    select(Year, Age, mx) %>%
    tidyr::pivot_wider(names_from = "Age", values_from = "mx") %>%
    remove_rownames %>% 
    column_to_rownames(var="Year") -> data
    
  data
}
rex <- function(mx, age){
  sum(cumprod(1-tail(mx, length(mx)-age)))  
}
full_rex <- function(lt)
{
  conv_to_A(lt) -> A
  data.matrix(A) -> A
  
  E <- matrix(0L, nrow = dim(A)[1], ncol = dim(A)[2])
  
  colnames(E) <- colnames(A)
  rownames(E) <- rownames(A)
  
  for (i in 1:nrow(A))
    {
    for(age in 0:ncol(A)-1)
      {
        E[i, age+1] <- rex(A[i,], age)
      }
    }
  E
}
```


```{r}
out_pivot <- purrr::map(out, conv_to_A)
out_rex <- purrr::map(out, full_rex)
out_pivot_log <- purrr::map(out_pivot, log)
#glimpse(out_pivot)
#glimpse(out_rex)
```


* **Visualization of residual life expectancy as a function of `Year` at ages $60$ and $65$, facetted by `Gender` and `Country` :**

```{r retirement, echo=FALSE, eval=FALSE}
ages <- c(60, 65)
re <- data_b_nl %>% 
  filter(as.numeric(Age) >= 60) %>%
  full_rex() %>%
  data.frame()
names(re) <-60:109
re %>% 
  rownames_to_column(var = "Year") -> re
re %>%
  pivot_longer(-Year, names_to = "Age", values_to = "rex") %>%
  filter(Age %in% ages) %>%
  ggplot(mapping = aes(x = Year, y = rex)) + 
  geom_point(aes(colour = Age)) + 
  scale_x_discrete(breaks = re$Year[seq(5,202, by = 50)]) +
  facet_wrap(~ Age) +
  geom_point(shape="+") +
  ggtitle('Residual life expectancy') +
  theme_bw()
```



# PCA  and SVD over log-mortality tables

- As we chose to study on the Netherlands as our European country, we also now pick a range of years `1948:2010`.
Then we extract the corresponding  lines from `life_table_pivot`, with taking logarithms of
central death rates. Once we did all that, we perform principal component analysis :


```{r}
plot_pca<-function(data, sca, cen){
  pca_data <- data
  
  pca_data[unlist(lapply(1948:2010, as.character)),] ->pca_data
  out_rex$data_b_nl[unlist(lapply(1948:2010, as.character)),1] -> rx
  pca_data <- cbind(pca_data, rx)
  res.pca = prcomp(pca_data, scale. = sca, center = cen)
  print(fviz_pca_ind(res.pca, repel = T))
  print(fviz_pca_var(res.pca,col.var = "contrib"))
  print(fviz_eig(res.pca))
}
plot_pca(out_pivot_log$data_f_nl, TRUE, TRUE)
#plot_pca(out_pivot_log$data_f_nl, FALSE, TRUE)
#plot_pca(out_pivot_log$data_f_nl, TRUE, FALSE)
#plot_pca(out_pivot_log$data_f_nl, FALSE, FALSE)
```


+ Concerning the screeplot :

Our scree plot displays how much variation each principal component captures from the data. Since our scree plot is a steep curve that bends quickly and flattens out, the first two PCs are sufficient to describe the essence of the data. So we can say that PCA works well on our data.

+ Concerning the correlation circle :

We see on the correlation circle that the infant mortality is inversely correlated with life expectancy. Indeed, all the advanced ages are tending down whereas the younger ages are tending up, on the left side of the circle. And the mx arrow is going to the right side of the circle. But we have to consider the fact that the oldest ages reprensent a small percentage of the total population.

+ Concerning the biplot :

We see that the recent years are more distributed on the right side of the biplot, which means that they follow the direction of mx on the correlation circle. So the PCA allows us to conclude that the life expectancy is getting higher as time goes by

# Canonical Correlation Analysis

+ We perform a Canonical Correspondance Analysis of $Z$ :

```{r}
cnty <- c("USA", "Netherlands")
life_table_pivot %>%
  filter(Country %in% cnty, Year %in% 1948:1998, Gender == 'Female' ) %>%
  select(-c("Gender","life_ex")) -> Z
us <-Z %>%
  filter(Country == "USA" ) %>%
  select(-c("Country")) %>%
  column_to_rownames(var="Year") %>%
  as.matrix()
nd <-Z %>%
  filter(Country == "Netherlands" ) %>%
  select(-c("Country")) %>%
  column_to_rownames(var="Year") %>%
  as.matrix()
n = names(us)
#names(us) <- c(n[1],paste(tail(n,length(n)-1) , "us", sep="-"))
#names(nd) <- c(n[1],paste(tail(n,length(n)-1) , "nd", sep="-"))
Z<-merge(nd,us)
print(Z)
#res.cc=cc(sp,nd)
#plt.cc(res.cc)
#cancor(sp, nd) -> cca
correl <- matcor(us, nd )
#img.matcor(correl, type = 2)
cca_2 <- cancor(nd,us)
#plot(cca_2)
#plt.cc(cca, var.label = TRUE, ind.names = data[,1])
#p.cca <- vegan::cca(Z)
#plot(p.cca, type ='t')
```

- [ ] Perform a Canonical Correspondance Analysis of $Z$



- [ ] Comment ...
- [ ] Comment ...

# Lee-Carter model for US mortality

During the last century, in the USA and in western Europe,
central death rates at all ages have exhibited a general decreasing trend.
This decreasing trend has not always been homogeneous across ages.
Governments and many stakeholders in the health and insurance sectors were intreseted in understanding these trends in order to more precisely model and predict the evolution of the mortality rate.  
One of the biggest hurdles was to correctly predict the age specific trends and variations of mortality.

The Lee-Carter model has been designed to model and forecast the
evolution of the log-central death rates for the United States during the XXth century.

The result is a matrix of age specific foracasted mortality rates.

Let $A_{x,t}$ denote the log central death rate at age $x$ during year $t\in T$
for a given population (defined by Gender and Country).

The Lee-Carter model assumes that observed loagrithmic central death rates
are sampled according to the following model
\[
A_{x,t} \sim_{\text{independent}} a_x + b_x \kappa_t + \epsilon_{x,t}
\]
where $(a_x)_x, (b_x)_x$ and $(\kappa_t)_t$ are unknown vectors that satisfy
\[
a_x = \frac{1}{|T|}\sum_{t \in T} A_{x,t}\qquad \sum_{t\in T} \kappa_t = 0 \qquad \sum_{x} b_x^2 =1
\]
and $\epsilon_{x,t}$ are i.i.d Gaussian random variables.

To estimate our vectors we have used the SVD decomposition  \[M = UDV^T\]

\[a_x\] and \[b_x = \frac{V_{x,1}}{\sum_{x} V_{.,1}} \] are the age dependent elements of our model and will be used for the rest of the modelisation.

\[k_t = D_1* U_{1,x} * \sum_{x} V_{.,1} \] is the time (Year) dependent element of the Lee-carter model, each value of this vector corresponds to a year.

The following graphs show that our Lee-Carter fit raging from 1933 to 1995 of US mortality data is acceptable and we can proceed with the prediction of our futur \[ k_t\] values.

## US data
- [ ] Fit a Lee-Carter model on the American data (for Male and Female data) training on years `1933` up to `1995`.
```{r}


Lee_carter_fit<- function(df, c, g, y_start, y_lc ,y_end){
  df %>% 
    filter(Country == c , Year %in% y_start:y_lc, Gender == g ) %>%
    select(-c(Gender,Country, Year)) -> m
  
  ax <- m %>%
    colMeans() %>% 
    as.matrix() 
  
  m %>%
    sweep(2, ax) %>%
    svd(1,1) -> df.svd
  
  #bx <- as.matrix(df.svd$v/sqrt(sum(df.svd$v**2)))
  bx <- as.matrix(df.svd$v/sum(df.svd$v))
  k_t <- t(sum(df.svd$v)*df.svd$d[1]*df.svd$u)

  drift_len = y_lc - y_start
  drift =(k_t[length(k_t)]-k_t[1])/length(k_t)
  for (i in 1:(y_end-y_lc)){
    eps  = rnorm(1, mean = drift, sd = sqrt(sd(k_t))/2)
    k_t <- cbind(k_t,k_t[length(k_t)] + eps)
  }
  
  lc <- sweep(bx %*% k_t, 1, ax, "+") %>%
    t() %>%
    as_tibble() 
  
  names(lc) <- 0:109
  lc$Year <- y_start:y_end
  lc
  }


```

``` {r }

plot_mortality_lee_carter <- function(df, c, g, y_start, y_lc ,y_end, years){
  
  lt_lc <- Lee_carter_fit(df, c, g, y_start, y_lc ,y_end)
  
  lt_lc <-lt_lc %>%
    pivot_longer(-Year,names_to = "Age", values_to = "mx_lc") %>%
    mutate(mx_lc = exp(mx_lc), Age = as.numeric(Age)) %>%
    filter(Year %in% years)
  
  real_lt <- life_table %>%
    filter(Country == c , Year %in% years, Gender == g ) %>%
    select(c(Year,Age,mx)) 
  
    
  lt_merged <- merge(real_lt, lt_lc)
  lt_merged <- lt_merged[order(lt_merged$Year, lt_merged$Age),]

  lc_plot <- lt_merged %>%
    select(c(Year,mx, mx_lc, Age)) %>%
    pivot_longer(-c(Year,Age),names_to = "method", values_to = "mx_n") %>%        
    ggplot(mapping=aes(x=Age, y=mx_n)) +
    geom_line(method="loess", se=FALSE, span=.5,aes(colour = forcats::as_factor(method)), size = .2) +
    scale_y_log10() +
    xlab("Age")+
    ylab("Mortality quotients") +
    labs(colour = "Mx method") +
    facet_wrap(~ Year) +
    ggtitle("Lee carter fit comparaison") +
    theme_bw() 
  print(lc_plot)
}

years <- seq(1935,1995, by = 20)

plot_mortality_lee_carter(life_table_pivot_log, 'USA', 'Male', 1933, 1995, 2040, years)

```



- [ ] Use the Lee-Carter model to predict the central death rates for years $2000$ up to $2015$

      
For the prediction we have used a random walk as mentionned in the 1992 publication by Ronald Lee:

\[k_{t} = k_{t-1} + d + \epsilon \] with d being the drift and epsilon a gaussian random variable with the same standard deviation as \[k_t\].
the drift is the trickiest part to estimate, it has a massive impact on the trend of ou prediction, we setteled on \[d = \frac{k_T - k_1}{T}\] 
which. We have used the same function as the one used for the fit and just concatenate the new \[k\] values to the fitted vector. the resulting matrix will have x lines corresponding to the fitted values and y lines for the predicted years.   


- [ ] Plot predictions and observations for years $2000, 2005, 2010, 2015$
```{r}

years <- seq(2000,2015, by = 5)
plot_mortality_lee_carter(life_table_pivot_log, 'USA', 'Male', 1933, 1995, 2040, years)

```


## Application of Lee-Carter model to a European Country

- [ ] Fit a Lee-Carter model to a European country
```{r}

years <- seq(1935,1995, by = 20)
plot_mortality_lee_carter(life_table_pivot_log, 'Netherlands', 'Male', 1933, 1995, 2040, years)
```
- [ ] Comment
The lee-Carter fit is very close to the observed values. Although it's smoother than the real mortality curve, it shouldn't impact our next prediction.

- [ ] Compare with rank-2 truncated SVD
- [ ] Use the Lee-Carter model to predict the central death rates for years $2000$ up to $2015$
      Plot predictions and observations for years $2000, 2005, 2010, 2015$

```{r}

years <- seq(2000,2015, by = 5)
plot_mortality_lee_carter(life_table_pivot_log, 'USA', 'Male', 1933, 1995, 2040, years)

```   
## Predictions of life expectancies at different ages

- [ ] Use Lee-Carter approximation to approximate residual life expectations
```{r}

get_lee_carter_life_exp <- function(df, c, g, y_start, y_lc ,y_end){
  
  new_lt <- Lee_carter_fit(df, c, g, y_start, y_lc ,y_end)
    
  new_lt <-new_lt %>%
    pivot_longer(-Year,names_to = "Age", values_to = "mx_lc") %>%
    mutate(mx_lc = exp(mx_lc), Age = as.numeric(Age))
  
  new_lt_pivot <- new_lt %>%
    select(Year, Age, mx_lc) %>%
    tidyr::pivot_wider(names_from = "Age", values_from = "mx_lc")
  
  new_lt_pivot %>%
      select(-Year) %>%
      apply(1, ex) -> new_lt_pivot$life_ex 
  
  select(new_lt_pivot, c(Year, life_ex))
}


lt_lc_us <- get_lee_carter_life_exp(life_table_pivot_log, 'USA', 'Female', 1933, 1995, 2040)


p <- lt_lc_us %>%
  ggplot(mapping=aes(x=Year, y=life_ex)) +
  geom_line(method="loess", se=FALSE, span=.5, size = .2) +
  scale_y_log10() +
  xlab("Year")+
  ylab("Life expectancy") +
  ggtitle("Lee carter fit comparaison") +
  theme_bw()

print(p)


```

```{r}


lee_carter_avg  <- function(df, c, g, y_start, y_lc ,y_end, nb){
  
  lc_le <- data.frame(Year = y_start:y_end )

  for (i in 1:nb){
    #tmp <- get_lee_carter_life_exp(df, c, g, y_start, y_lc ,y_end)
    lc_le[i+1] <- get_lee_carter_life_exp(df, c, g, y_start, y_lc ,y_end)['life_ex']
  }
  lc_le %>%
    select(-Year) %>%
    rowMeans() -> lc_le$Mean
  lc_le %>%
  select(-c(Year, Mean)) %>%
  apply(1,quantile,probs = c(10,90)/100) -> quantiles
  
  lc_le$quantile_90 <- quantiles[2,] 
  lc_le$quantile_10 <- quantiles[1,] 
  lc_le$Country <- c
  #lc_le$Mean[1:y_start-y_lc+1] <- filter(life_table_pivot, Country == c , Year %in% y_start:y_lc, Gender == g)$life_ex
  lc_le %>% 
    select(c(Year, Mean, quantile_10, quantile_90, Country)) %>%
    filter(Year %in% y_lc:y_end)
}


y_start = 1945
y_lc = 1995
y_end = 2050
g = "Both"
us_le_lc <- lee_carter_avg(life_table_pivot_log, 'USA', g,  y_start, y_lc ,y_end, 50)
nl_le_lc <- lee_carter_avg(life_table_pivot_log, 'Netherlands', g,  y_start, y_lc ,y_end, 50)
sw_le_lc <- lee_carter_avg(life_table_pivot_log, 'Sweden', g,  y_start, y_lc ,y_end, 50)


life_table_pivot %>%
  filter(Country %in% c("USA","Netherlands") , Year %in% y_start:2015, Gender == g) %>%
  select(c(Year, Country,life_ex)) %>%
  pivot_longer(-c(Year,Country),names_to = "Percentile", values_to = "life_exp") -> real_lexp
```

``` {r}

p <- rbind(us_le_lc, nl_le_lc) %>%
  select(c(Year,Mean,quantile_10,quantile_90,Country )) %>%
  pivot_longer(-c(Year,Country),names_to = "Percentile", values_to = "life_exp") %>%
  rbind(real_lexp) %>%
  ggplot() +
  geom_line(aes(x=Year, y=life_exp,linetype=forcats::as_factor(Percentile), colour = forcats::as_factor(Country)),method="loess", se=FALSE, span=.5, size = .2) +
  #geom_line(data=real_lexp, map=aes(x=Year, y=life_ex, colour = forcats::as_factor(Country)),method="loess", se=FALSE, span=.5, size = .2) +
  scale_y_log10() +
  xlab("Year")+
  ylab("Life expectancy") +
  labs(colour = "Mx method") +
  ggtitle("Lee carter fit comparaison") +
  theme_bw()

print(p)

  
```

- [ ] Compare with observed residual life expectations

# References

__Life tables and demography__

- [Human Mortality Database](https://www.mortality.org)
- [Tables de mortalité françaises, Jacques Vallin et France Meslé](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf)
- [Modeling and Forecasting U.S. Mortality, R.D.Lee and L.R. Carter, JASA 1992]
- [Les dimensions de la mortalité, S. Ledermann, Jean Breas, Population, 1959]

__Graphics and reporting__

- [Interactive web-based data visualization with R, plotly, and shiny](https://plotly-r.com/index.html)
- [R for Data Science](https://r4ds.had.co.nz)
- [Layered graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)
- [Plotly](http://plotly.com/)

__Tidyverse__

- [tidyselect](https://tidyselect.r-lib.org/articles/tidyselect.html)
- [dbplyr](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html)
- [data.table](https://github.com/Rdatatable/data.table)
- [DT](https://rstudio.github.io/DT/)

__PCA, SVD, CCA__

- [FactoMineR](http://factominer.free.fr/index_fr.html)
- [ade4](http://pbil.univ-lyon1.fr/ade4/accueil.php)
- [FactoInvestigate](http://factominer.free.fr/reporting/index_fr.html)
- [PCA and Tidyverse](https://cmdlinetips.com/2019/05/how-to-do-pca-in-tidyverse-framework/)
- [tidyprcomp](https://broom.tidyverse.org/reference/tidy.prcomp.html)