---
title: 'Homework II: SVD analysis & Life Tables'
subtitle: 'Deadline: 2020-05-20'
author: "Inès Dardouri, Mohamed Boudokhane and Lycia Fezzoua"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    number_sections: yes
    toc: yes
  html_notebook:
    code_folding: none
    number_sections: yes
    toc: yes
params:
    dirpath: '../LIFE_TABLES/lt_'
    country_code: 'fr_t'
    country: 'France'
    timecourse: 1945:2015
---


---

# Objectives

This notebook aims at

  - working with **tables** (`data.frames`, `tibbles`, `data.tables`, ...) using `dplyr` or any other query language (as provided for example by `data.table`)
  - visualizing demographic data as provided by [Human Mortality Database organization](https://www.mortality.org).
  - using PCA and other matrix oriented methods (CCA) to explore a multivariate datasets (lifetables may be considered as multivariate datasets)

# Life tables data (ETL)

Life data tables are downloaded from [https://www.mortality.org](https://www.mortality.org).

See also [https://www.lifetable.de](https://www.lifetable.de).

If you install and load package [https://cran.r-project.org/web/packages/demography/index.html](https://cran.r-project.org/web/packages/demography/index.html), you will also find life data tables.

We investigate life tables describing countries from Western Europe (France, Great Britain --actually England and Wales--, Italy, the Netherlands, Spain, and Sweden) and the United States.

We load the one-year lifetables for female, male and whole population for the different countries.

```{r, echo=FALSE, eval=FALSE}
# for debugging
# params should be initialized from YAML header
params<- list(
    timecourse= 1945:2015,
    dirpath= '../LIFE_TABLES/lt_',
    country_code= 'fr_t',
    country= 'France')
```

```{r, echo=FALSE}
timecourse <- eval(rlang::parse_expr(params$timecourse))
```

```{r tidyverse, message=FALSE, warning=FALSE, echo=FALSE}
pacman::p_load(tidyverse)
pacman::p_load(plotly)
pacman::p_load(foreach)
pacman::p_load(iterators)
pacman::p_load(DT)
pacman::p_load(ade4)
pacman::p_load(FactoMineR)
pacman::p_load(factoextra)
pacman::p_load(FactoInvestigate)
pacman::p_load(ggfortify)

old_theme <-theme_set(theme_dark(base_size=9,
                                 base_family = "Helvetica"))

knitr::opts_chunk$set(eval=FALSE,
  echo=FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  autodep = TRUE,
  tidy = FALSE)

```


```{r, echo=FALSE}
country_code <- list(fr_t='FRATNP',
                     fr_c='FRACNP',
                     be='BEL',
                     gb_t='GBRTENW',
                     gb_c='GBRCENW',
                     nl='NLD',
                     it='ITA',
                     swe='SWE',
                     sp='ESP',
                     us='USA')

countries <- c('fr_t',  'gb_t',  'nl', 'it', 'sp', 'swe', 'us')

country_names <- list(fr_t='France',     # total population
                     fr_c='France',      # civilian population
                     be='Belgium',
                     gb_t='England & Wales',    # total population
                     gb_c='England & Wales',    # civilian population
                     nl='Netherlands',
                     it='Italy',
                     swe='Sweden',
                     sp='Spain',
                     us='USA')

gender_names <- list('b'='Both',
                     'f'='Female',
                     'm'='Male')
```


Check on [http://www.mortality.org](http://www.mortality.org) the meaning of the different columns:

Document [Tables de mortalité françaises pour les XIXe et XXe siècles et projections pour le XXIe siècle](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf) contains
detailed information on the construction of Life Tables for France.
Two kinds of Life Tables can be distinguished: *Table du moment* which
contain for each calendar year, the mortality risks at different ages for that very year; and *Tables de génération* which contain for a given birthyear, the   mortality risks
at which an individual born during that year has been exposed. The life tables investigated in this homework are *Table du moment*. According to the document by Vallin and Meslé, building the life tables  required decisions and doctoring.

See (among other things)

- p. 19 Abrupt changes in mortality quotients at some ages for a given calendar year
- Estimating mortality quotients at great age.

Have a look at [Lexis diagram](https://en.wikipedia.org/wiki/Lexis_diagram).

Some columns need retyping:

- `Year`: should be integer
- `Age`:  needs some cleaning, after cleaning it should be typed as integer
- `Lx`: should be integer
- `Tx`: should be integer
- Other columns should be considered as floating point numbers (`numeric`)

---

- [ ] After ETL processing, you obtain a universal table encompassing
all data available  in the files located in `LIFE_TABLES` directory.
Henceforth, the universal table is named `life_table`, its schema should be
the following.

| Column Name | Column Type |
|:------------|:------------|
|  Year       | integer     |
|  Age        | integer     |
|  mx         | double      |
|  qx         | double      |
|  ax         | double      |
|  lx         | integer     |
|  dx         | integer     |
|  Lx         | integer     |
|  Tx         | integer     |
|  ex         | double      |
|  Country    | factor      |
|  Gender     | factor      |

Coercion introduces a subtantial number of NA  warnings.
Preliminary inspection of the data suggests that coercion problems
orginate from column `Age`: `110+` cannot be coerced to an integer value.
Discard corresponding rows using `tidyr::drop_na(Age)`.


# Western countries in 1948

- [ ] Plot the central death rates of all Countries at all ages for year 1948.
- [ ] Comment.
- [ ] Plot ratios between central death rates in European countries and
central death rates in the USA in 1948.
- [ ] Comment

# Death rates evolution since WW II

- [ ] Plot mortality quotients (column `mx`) for
both genders as a function of `Age` for  years `1946, 1956, ...` up to `2016` .
Use aesthetics to distinguish years.

- [ ] Facet by `Gender`  and `Country`

- [ ] Write a function `ratio_mortality_rates` with signature
`function(df, reference_year=1946, target_years=seq(1946, 2016, 10))`
that takes as input:

  - a dataframe with the same schema as `life_table`,
  - a reference year `ref_year` and
  - a sequence of years `target_years`

and that returns a dataframe with schema:


| Column Name | Column Type |
|:------------|:------------|
|  Year       | integer     |
|  Age        | integer     |
|  mx         | double      |
|  mx.ref_year| double      |
|  Country    | factor      |
|  Gender     | factor      |

where `(Country, Year, Age, Gender)` serves as a _primary key_,
`mx` denotes the central death rate at `Age` for `Year` and `Gender` in `Country`
whereas `mx_ref_year` denotes central death rate at `Age` for argument `reference_year`
in `Country` for `Gender`.


```{r}
ratio_mortality_rates <- function(df,
                                  reference_year=1946,
                                  target_years=seq(1946, 2016, 10)){

}
```


- [ ] Draw plots displaying the ratio  $m_{x,t}/m_{x, 1946}$ for ages $x \in 1, \ldots, 90$
and year $t$ for $t \in 1946, \ldots, 2016$ where $m_{x,t}$ is the central death rate
at age $x$ during year  $t$.

  1. Handle both genders and countries `Spain`, `Italy`, `France`, `England & Wales`, `USA`, `Sweden`, `Netherlands`.
  1. One properly facetted plot is enough.

- [ ] Comment


# Trends


- [ ] Plot mortality quotients at ages $0, 1, 5$  as a function of time. Facet by Gender and Country

- [ ] Comment

- [ ] Plot mortality quotients at ages $15, 20, 40, 60$  as a function of time. Facet by `Gender` and `Country`

- [ ] Comment

# Rearrangement

- [ ] From dataframe `life_table`, compute another dataframe  called `life_table_pivot`
with primary key `Country`, `Gender` and `Year`, with a column for each `Age` from `0` up to `110`.
For each age column, the entry should be the central death rate
at the age defined by column, for `Country`, `Gender`  and `Year` identifying the row.

You may use functions `pivot_wider`, `pivot_longer` from `tidyr::` package.

The resulting schema should look like:

| Column Name | Type    |
|:------------|:--------|
| Country     | factor  |
| Gender      | factor  |
| Year        | integer |
| `0`         | double  |
| `1`         | double  |
| `2`         | double  |
| `3`         | double  |
| $\vdots$    | $\vdots$|

- [ ] Using `life_table_pivot` compute life expectancy at birth for each Country, Gender and Year

# Life expectancy

- [ ] Write a function that takes as input a vector of mortality quotients, as well as an age, and returns the residual life expectancy corresponding to the vector and the given age.

- [ ] Write a function that takes as input
a dataframe with the same schema as  `life_table` and returns
a data frame with columns `Country`, `Gender`, `Year`, `Age` defining a primary key
and a column `res_lex` containing _residual life expectancy_ corresponding to the pimary key.

In order to compute residual life expectancies, you may consider using `window` functions
oer apropriately defined windows. Package `dplyr` does not offer a rich API for window functions.
Package `dbplyr` does.

- [ ] Plot residual life expectancy as a function of `Year` at ages $60$ and $65$,
facet by `Gender` and `Country`.

```{r retirement, echo=FALSE, eval=FALSE}
```



# PCA  and SVD over log-mortality tables

- [ ] Pick a Country, a Gender,  a range of years `1948:2010`.
Extract the corresponding  lines from `life_table_pivot`. Take logarithms of
central death rates and perform principal component analysis.
- [ ] comment the screeplot
- [ ] comment the correlation circle
- [ ] comment the biplot

# Canonical Correlation Analysis

- [ ] Build a function that takes as input
  + a dataframe like `life_table_pivot`,
  + a couple  of countries, say `Spain` and `Sweeden`,
  + a vector of `Year`, say `1948:1998`
  + a `Gender` say `Female`
  returns a matrix called $Z$ with rows corresponding to `Year` and columns
  corresponding to couples `(Country, Age)`.

- [ ] Perform a Canonical Correspondance Analysis of $Z$
- [ ] Comment ...
- [ ] Comment ...

# Lee-Carter model for US mortality

During the last century, in the USA and in western Europe,
central death rates at all ages have exhibited a general decreasing trend.
This decreasing trend has not always been homogeneous across ages.

The Lee-Carter model has been designed to model and forecast the
evolution of the log-central death rates for the United States during the XXth century.

Let $A_{x,t}$ denote the log central death rate at age $x$ during year $t\in T$
for a given population (defined by Gender and Country).

The Lee-Carter model assumes that observed loagrithmic central death rates
are sampled according to the following model
\[
A_{x,t} \sim_{\text{independent}} a_x + b_x \kappa_t + \epsilon_{x,t}
\]
where $(a_x)_x, (b_x)_x$ and $(\kappa_t)_t$ are unknown vectors that satisfy
\[
a_x = \frac{1}{|T|}\sum_{t \in T} A_{x,t}\qquad \sum_{t\in T} \kappa_t = 0 \qquad \sum_{x} b_x^2 =1
\]
and $\epsilon_{x,t}$ are i.i.d Gaussian random variables.

## US data

- [ ] Fit a Lee-Carter model on the American data (for Male and Female data) training on years `1933` up to `1995`.
- [ ] Compare the fit provided by the Lee-Carter model with the fit provided by a rank $2$
truncated SVD
- [ ] Compare vectors avec $(a_x)_x, (b_x)_x$ and $(\kappa_t)_t$ with appropriate singular vectors.
- [ ] Use the Lee-Carter model to predict the central death rates for years $2000$ up to $2015$
- [ ] Plot predictions and observations for years $2000, 2005, 2010, 2015$

## Application of Lee-Carter model to a European Country

- [ ] Fit a Lee-Carter model to a European country
- [ ] Comment
- [ ] Compare with rank-2 truncated SVD
- [ ] Use the Lee-Carter model to predict the central death rates for years $2000$ up to $2015$
      Plot predictions and observations for years $2000, 2005, 2010, 2015$

## Predictions of life expectancies at different ages

- [ ] Use Lee-Carter approximation to approximate residual life expectations
- [ ] Compare with observed residual life expectations

# References

__Life tables and demography__

- [Human Mortality Database](https://www.mortality.org)
- [Tables de mortalité françaises, Jacques Vallin et France Meslé](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf)
- [Modeling and Forecasting U.S. Mortality, R.D.Lee and L.R. Carter, JASA 1992]
- [Les dimensions de la mortalité, S. Ledermann, Jean Breas, Population, 1959]

__Graphics and reporting__

- [Interactive web-based data visualization with R, plotly, and shiny](https://plotly-r.com/index.html)
- [R for Data Science](https://r4ds.had.co.nz)
- [Layered graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)
- [Plotly](http://plotly.com/)

__Tidyverse__

- [tidyselect](https://tidyselect.r-lib.org/articles/tidyselect.html)
- [dbplyr](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html)
- [data.table](https://github.com/Rdatatable/data.table)
- [DT](https://rstudio.github.io/DT/)

__PCA, SVD, CCA__

- [FactoMineR](http://factominer.free.fr/index_fr.html)
- [ade4](http://pbil.univ-lyon1.fr/ade4/accueil.php)
- [FactoInvestigate](http://factominer.free.fr/reporting/index_fr.html)
- [PCA and Tidyverse](https://cmdlinetips.com/2019/05/how-to-do-pca-in-tidyverse-framework/)
- [tidyprcomp](https://broom.tidyverse.org/reference/tidy.prcomp.html)
