---
title: "Homework I: Life Tables"
subtitle: "Cours: Analyse de Données M1 Mathématiques/M1 Mathématiques-Informatique"
author: "Inès Dardouri et Mohamed Boudokhane"
output:
  html_notebook:
    code_folding: none
    number_sections: yes
    toc: yes
---

# Objectives

This notebook aims at

- working with **tables** (`data.frames`, `tibbles`, `data.tables`, ...) using `dplyr` or any other query language (as provided for example by `data.table`)
- visualizing demographic data as provided by [Human Mortality Database organization](https://www.mortality.org).# Life dataLife data tables can be downloaded from [https://www.mortality.org](https://www.mortality.org). Some data have been downloaded for you and are available [here]().

See also [https://www.lifetable.de](https://www.lifetable.de).

If you install and load package [https://cran.r-project.org/web/packages/demography/index.html](https://cran.r-project.org/web/packages/demography/index.html), you will also find life data tables.

```{r}
pacman::p_load(tidyverse)
pacman::p_load(tidymodels)
```

## Life table inspection

Load the data corresponding to the country you are supposed to work on.You may use `readr::read_delim()` with appropriate options.

For country denoted by country code `xxx`, you will load three life tables:

- `xxx.fltper_1x1.txt` where `f` stands for _female_
- `xxx.mltper_1x1.txt` where `m` stands for _male_
- `xxx.bltper_1x1.txt` where `b` stands for _both_

Name the data.frames `data_f`, `data_m`, `data_b`.

```{r}
country_code <- list(fr='FRACNP',
                     be='BEL',
                     gb='GBRTENW',
                     nl='NLD',
                     it='ITA',
                     swe='SWE',
                     sp='ESP',
                     us='USA')

country <- 'fr'

ww_I <- 1913:1921
#ww_I <- 1938:1946 # (US)
```

```{r}
# TODO:
data_f = readr::read_delim(paste("LIFE_TABLES/",country_code[country],".fltper_1x1.txt",sep=""), delim=" ", skip = 2)
data_m = readr::read_delim(paste("LIFE_TABLES/",country_code[country],".mltper_1x1.txt",sep=""), delim=" ", skip = 2)
data_b = readr::read_delim(paste("LIFE_TABLES/",country_code[country],".bltper_1x1.txt",sep=""), delim=" ", skip = 2)

```

For each life table, each row is identified by `Year` and `Age`.
Read the documentation and concisely explain the meaning of columns

Check on [http://www.mortality.org](http://www.mortality.org) the meaning of the different columns:

`mx`
: Central death rate between ages x and x+n where n=1, 4, 5, or ∞ (open age interval)

`qx`
: Probability of death between ages x and x+n

`ax`
: Average length of survival between ages x and x+n for persons dying in the interval

`lx`
: Number of survivors at exact age x, assuming l(0) = 100,000

`dx`
: Number of deaths between ages x and x+n

`ex`:
: Life expectancy at exact age x (in years)

- What is the class of `data_b, ...`?
- Inspect `data_f` using `glimpse`.

```{r}
# TODO:
#class(data_b)
glimpse(data_f)

v=c()
for (i in 1:ncol(data_b)){
  v=c(v,(is.na(data_b[,i])))
}
v

```

Which columns needs retyping?
: TODO:

## Life table cleaning

Remove white spaces from column names. You may find useful functions in package `stringr`.
Try to follow the **DRY** principle (Don't Repeat Yourself).
You need to perform the same operation on every column name.

You may use fonctional programming (using `purrr::` or `plyr::`)
to process all column names of `data_xxx` using one expression.
Some verbs in `dplyr::` provide you with shortcuts.Write a function `retype` that can tidy data.frames
with the same schema as `data_xxx` (cleaning column names if necessary, casting
some columns to `numeric`, others to `double`).

```{r}
retype <- function(data){
  names(data) %>%
  stringr::str_replace_all(., ' ', '') -> names(data)
  data %>%
    dplyr::mutate_at(vars(mx, qx, ax, ex), as.double) %>%
    dplyr::mutate_at(vars(Year, lx, dx, Lx, Tx), as.integer) %>%
    dplyr::mutate_at(vars(Age), as.factor) -> data
  data
}

glimpse(data_f)
```

Useful tools:

- `dplyr::mutate_at`
- `dplyr::rename`
- `dplyr::rename`Apply function `retype` to the three data.frames.

```{r}
# TODO:

count_na <- function(data){
  data %>%
    dplyr::summarise_all(.funs = ~sum(is.na(.)))-> res
  res
}

retype(data_f) -> data_f
retype(data_m) -> data_m
retype(data_b) -> data_b

count_na(data_f)

```

# Hazard rates, mortality quotients

Plot mortality quotients (column `mx`) for women as a function of `Age` for years `1913` up to `1921`.
Use aesthetics to distinguish years. You will need to categorize the `Year` column
(`forcats::` may be helpful).

Pay attention to axes labels, to legends. Assess logarithmic scales.

```{r}
# TODO:
pacman::p_load(viridis)
p <- data_f %>%
  filter(Year %in% ww_I) %>%
  ggplot(mapping=aes(x=Age, y=mx)) +
  geom_point(aes(colour = forcats::as_factor(Year)), size = .5) +
  scale_x_discrete(breaks = data_f$Age[c(TRUE,FALSE,FALSE)]) +
  #facet_grid(~ Year) +
  xlab("Age")+
  ylab("mortality quotients for women") +
  labs(colour = "Years")

print(p)
```

Plot mortality quotients (column `mx`) for men as a function of `Age` for years `1913` up to `1921`.

Note that you may tweak the `ggplot` object you just created and feed it with new data
using `%+%` from `ggplot`.

```{r}
# TODO:
p %+%
  filter(data_m, Year %in% ww_I)
```

## Facetting

Build a data.frame called `big_data`, where rows are
identified by `Year, Age, Sex` and comes either
from `data_f` or `data_m`. Column `Sex` should be a `factor`
with two levels called `Male` and `Female`.

You may proceed as follows:- Add a character column "Sex" to `data_f` and `data_m`.

- Stack the two data.frames (using `rbind`) into data.frame named `big_data`.
- Cast column `Sex` as a factor.

```{r}
# TODO:
data_m$Sex <- "male"
data_f$Sex <- "female"
big_data = rbind(data_f, data_m)
big_data%>%
  dplyr::mutate(Sex = forcats::as_factor(Sex)) -> big_data

glimpse(big_data)
```

Plot mortality quotients (column `mx`) as a function of `Age` for
years `1913` up to `1921` facet according to `Sex`.

```{r}
# TODO:
big_data %>%
  filter(Year %in% ww_I) %>%
  ggplot(mapping=aes(x=Age, y=mx)) +
  geom_point(aes(colour = Sex), shape = '+', size = 2) +
  scale_x_discrete(breaks = big_data$Age[c(TRUE, FALSE, FALSE)]) +
  #facet_grid(~ Year) +
  xlab("Age")+
  ylab("mortality quotients") +
  labs(colour = "Sex")

```

Map `Sex` to some aesthetics and facet by `Year`.

```{r}
# TODO:
```

List puzzling features on those plots.
: TODO:

Do you have clues about those puzzling aspects?
: TODO:# Trends## Infant mortality ratePlot mortality quotients at ages $0, 1, 5$ as a function of time.
Distinguish women and men.

```{r}
ages <- c(0, 1, 5)
# TODO:
ages
mort_by_age <- big_data %>%
  filter(Year %in% ww_I, as.numeric(levels(Age)) %in% ages) %>%
  ggplot(mapping=aes(x=Year, y=mx)) +
  geom_point(aes(colour = Sex, shape = Age) , size = 2)
mort_by_age

```

Can you match abrupt changes and historical events?
: TODO:

## As times goes byPlot mortality quotients at ages $15, 20, 40, 60$ as a function of time

Distinguish women and men.

Note again that you may tweak the `ggplot` object you just created and feed it with new data
using `%+%` from `ggplot`.

```{r}
ages <- c(15, 20, 40, 60)

# TODO:
mort_by_age %+%
    filter(big_data, Year %in% ww_I, as.numeric(levels(Age)) %in% ages)
```

TODO:
: List puzzling features

## Mortality quotients

- Plot mortality quotient as a function of age for years $1825, 1850, 1875, \ldots, 2000$, facet with respect to `Sex`
- Compare different scales (logarithmic or linear) on the different axes

```{r}
years <- seq(1825, 2000, by = 25)

# TODO:

p %+%
  filter(big_data, Year %in% years)

#logartithmic
```

# Rearrangement

From data.frame `data_m, data_f`, extract a numerical matrix `A` where rows correspond to year (say `t`), columns to age (say `x`), and entry
for year `t` and age `x` is the mortality quotient at age `x` for year `t`. Finally, we take logarithms: $A_{t,x} = \log m_{t,x}$.

You may use functions `pivot_wider`, `pivot_longer` from `tidyr::` package.

```{r}
# TODO:

```

> Factorizations of $A$ play a significant role in demographic modeling. More on this topic in PCA chapter.

# Life expectancy

Write a function that takes as input a vector of mortality quotients, as well as an age, and returns the residual life expectancy corresponding to the vector and the given age.

```{r}
# TODO:
```

```{r}
rex <- function(mx, age){
  # TODO:
}

```

Write a function that takes as input
a _life table_ and returns a matrix $E$ where rows correspond to `Year`
and columns to `Age` with $E_{t,x}$ equal to the residual life expectancy at age $x$ for year $t$.

```{r}
# TODO:
full_rex <- function(lt)
{
  # TODO:
}
```

Plot residual life expectancy as a function of `Year` at ages $60$ and $65$.
You may proceed by first performing aggregation on `data_...` (grouping by `Year`),
and then compute residual life expectancy at different ages
by slicing the `mx` column. Recall that when mapping hazard rate to
residual life expextancy, the order of multiplications matters. How could you secure
your computation so that it still works if the rows of `data_...` are shuffled ?

Function `tidyr::pivot_longer` may help.

```{r}
# TODO:
```

# References

- [Human Mortality Database](https://www.mortality.org)
- [Tables de mortalité françaises, Jacques Vallin et France Meslé](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf)
